---
title: "Building Your Own Custom Pipeline Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Your Own Custom Pipeline Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One key strength of the preprocessing framework within `eyeris` is its modularity. 

While we encourage most users to use the `glassbox()` function for simplicity and reproducibility, advanced users can create custom preprocessing steps that 
seamlessly integrate into the pipeline.

This vignette walks you through the structure required to write your own 
`eyeris`-compatible preprocessing functions.

## 🧩 How the Pipeline Works

Under the hood, each preprocessing function in `eyeris` is a wrapper around a 
core operation that gets tracked, versioned, and stored using the 
`pipeline_handler()`. 

Custom pipeline steps must conform to the `eyeris`
protocol for maximum compatibility with the downstream functions we provide.

Following the `eyeris` protocol also ensures:
- all operations follow a predictable structure, and
- that new pupil data columns based on previous operations in the chain are able
to be dynamically constructed within the core `timeseries` data frame.

For instance: 
`pupil_raw -> pupil_raw_deblink -> pupil_raw_deblink_detransient -> ...`

If you're unfamiliar with how these columns are structured and tracked, 
first check out the companion vignette: 
[📦 Anatomy of an `eyeris` Object](anatomy.html).

## 🛠️Creating a Custom Extension for `eyeris`

Let's say you want to write a new `eyeris` extension function called `winsorize()`
  to apply [winsorization](https://en.wikipedia.org/wiki/Winsorizing) to extreme
  pupil values.
  
### 1) Write the core operation function

This function should accept a data frame `x`, a string `prev_op` (i.e., the name
of the previous pupil column), and any custom parameters.

#### To illustrate:

```{r eval=FALSE}
winsorize_pupil <- function(x, prev_op, lower = 0.01, upper = 0.99) {
  vec <- x[[prev_op]]
  q <- quantile(vec, probs = c(lower, upper), na.rm = TRUE)
  vec[vec < q[1]] <- q[1]
  vec[vec > q[2]] <- q[2]
  vec
}
```

### 2) Create the wrapper using the `eyeris::pipeline_handler()`

The `pipeline_handler()` enables your function to automatically:

- track your function within the `eyeris` list object's `params` field,
- append a new column to each block within the `timeseries` list of data frames, and
- update the object's `latest` pointer.

#### To illustrate:

```{r eval=FALSE}
#' Winsorize pupil values
#'
#' Applies winsorization to extreme pupil values within each block.
#'
#' @param eyeris An `eyeris` object created by [load_asc()].
#' @param lower Lower quantile threshold. Default is 0.01.
#' @param upper Upper quantile threshold. Default is 0.99.
#' @param call_info A list of call information and parameters. If not provided,
#'   it will be generated from the function call.
#'
#' @return Updated `eyeris` object with new winsorized pupil column.
winsorize <- function(eyeris, lower = 0.01, upper = 0.99, call_info = NULL) {
  # create call_info if not provided
  call_info <- if (is.null(call_info)) {
    list(
      call_stack = match.call(),
      parameters = list(lower = lower, upper = upper)
    )
  } else {
    call_info
  }

  # handle binocular objects
  if (is_binocular_object(eyeris)) {
    # process left and right eyes independently
    left_result <- eyeris$left |>
      pipeline_handler(
        winsorize_pupil,
        "winsorize",
        lower = lower,
        upper = upper,
        call_info = call_info
      )
    
    right_result <- eyeris$right |>
      pipeline_handler(
        winsorize_pupil,
        "winsorize",
        lower = lower,
        upper = upper,
        call_info = call_info
      )
    
    # return combined structure
    list_out <- list(
      left = left_result,
      right = right_result,
      original_file = eyeris$original_file,
      raw_binocular_object = eyeris$raw_binocular_object
    )

    class(list_out) <- "eyeris"

    return(list_out)
  } else {
    # regular eyeris object, process normally
    eyeris |>
      pipeline_handler(
        winsorize_pupil,
        "winsorize",
        lower = lower,
        upper = upper,
        call_info = call_info
      )
  }
}
```

### 3) Understanding Call Stack Tracking

The `call_info` parameter is crucial for maintaining reproducibility and debugging in `eyeris`. It captures:

- **`call_stack`**: The exact function call that was made (using `match.call()`)
- **`parameters`**: A list of all parameters passed to your function

This information is automatically stored in `eyeris$params[[new_suffix]]` and can be used for:
- Generating reproducible reports
- Debugging pipeline issues
- Tracking parameter changes across different runs

**Important Notes:**
- Always include `call_info = NULL` as the last parameter in your function signature
- Use `match.call()` to capture the function call
- Pass all your function parameters in the `parameters` list
- Always pass `call_info = call_info` as the last argument to `pipeline_handler()`

### 4) Function Structure Breakdown

1. **Function signature**: Always include `call_info = NULL` as the last parameter
2. **Call info creation**: Create the call_info list if not provided
3. **Pipeline handler call**: Pass all parameters including call_info
4. **Parameter order**: 
   - First: `eyeris` object
   - Second: your core operation function name
   - Third: the suffix for column naming
   - Fourth+: your function parameters
   - Last: `call_info = call_info`
   
## 🎉 And that's it!

You should now be able to use your new function extension as a component within 
a new custom `eyeris` pipeline declaration. To illustrate:

```{r eval=FALSE}
system.file("extdata", "memory.asc", package = "eyeris") |>
  eyeris::load_asc(block = "auto") |>
  eyeris::deblink(extend = 50) |>
  winsorize()
```

## 💪 Best Practices

- **Use consistent naming:** i.e., match your suffix (e.g. `"winsorize"`) to the 
column and log structure:
  - If you call `pipeline_handler(..., "winsorize")`
  - Then, your function names should reflect that:
    - the public facing `winsorize()` wrapper function, and
    - the private `winsorize_pupil()` logic implementation function.
  - You will then see `pupil_raw_*_winsorize` as the new output column!
- **Respect previous steps:** Your custom function should rely only on `prev_op`,
not on **_any hardcoded column names_**!
- **Return the expected data type:** Be sure that you private function always 
returns a modified vector type, as the underlying `pipeline_handler()` is
looking out for a vector it can transpose into the new column that will be added
to the `timeseries` data frame within the resulting `eyeris` object.
- **Test on individual blocks:** Always try your private function logic on a 
single data block (e.g., `eyeris$timeseries[[1]]`) to debug before integrating
it with the `eyeris` pipeline protocol.
- **Include call stack tracking:** Always implement proper `call_info` handling for
reproducibility and debugging capabilities.

## 🔍 Advanced: Custom Call Info Handling

For more complex functions, you might want to customize the `call_info` structure:

```{r eval=FALSE}
custom_function <- function(eyeris, param1, param2, call_info = NULL) {
  # custom call_info with additional metadata
  call_info <- if (is.null(call_info)) {
    list(
      call_stack = match.call(),
      parameters = list(param1 = param1, param2 = param2),
      metadata = list(
        timestamp = Sys.time(),
        version = "1.0.0",
        description = "Custom processing step"
      )
    )
  } else {
    call_info
  }

  pipeline_handler(
    eyeris,
    custom_pupil_function,
    "custom",
    param1 = param1,
    param2 = param2,
    call_info = call_info
  )
}
```

## ✨ Summary

We hope you are now convinced at the power and extensibility the `eyeris`
protocol enables! As we demonstrated here, with just a little bit of structure,
you can create custom extension steps tailored to your specific analysis
needs -- all while preserving the reproducibility and organizational core
principles `eyeris` was designed and built around.

If you'd like to contribute new steps to `eyeris`, feel free to open a pull 
request or discussion on [GitHub](https://github.com/shawntz/eyeris)!

---

## 📚 Citing `eyeris`

<div class="alert alert-light" style="padding-bottom: 0;">
  If you use the `eyeris` package in your research, please cite it!
  
  Run the following in R to get the citation:
</div>

```{r}
citation("eyeris")
```
